\MakeUppercase{Layered $H$-Partitions of $k$-Planar Graphs} \thanks{This work was partly funded by NSERC and MRI.}

### 2019-05-14 k-planar.tex
First crack at (g,k)-planar graphs

\begin{abstract}
  Layered $H$-partitions of graphs (with small layered width, in which $H$ has small treewidth) are a recently-introduced tool that have been used to solve longstanding problems on queue layouts, non-repetitive colouring, and 3-d graph drawing.  Such partitions are known to exist for planar graphs and, more generally, bounded genus graphs.  In the current paper, we prove that every $k$-planar graph has a layered $H$-partition of layered width $O(k^2)$ in which $H$ has treewidth $O(c^k)$, for some constant $c$. This is the first result of this type for a non-minor-closed class of graphs and implies that $k$-planar graphs have both queue number and non-repetitive chromatic number upper-bounded by a function of $k$.  The former result was previously shown using a combination of $H$-partitions for planar graphs with \textit{ad hoc} methods. The latter result is new, for all $k\ge 1$.
\end{abstract}

@@

In this paper we show that another family of graphs admits layered $H$-partitions of constant layered-width in which $H$ has constant treewidth: $k$-planar graphs, which we now define. An \emph{embedded graph} $G$ is a graph with $V(G)\subset\R^2$ in which each edge $vw\in E(G)$ is a closed curve\footnote{A closed curve is a continuous function $f:[0,1]\to \R^2$. The points $f(0)$ and $f(1)$ are called the \emph{endpoints} of the curve.  When there is no danger of misunderstanding we treat a curve $f$ as a point set $\{f(t):0\le t\le 1\}$.} with endpoints $v$ and $w$ and not containing any vertex of $G$ in its interior.  A \emph{crossing} in an embedded graph $G$ is a triple $(p,vw,xy)$ with $p\in\R^2$, $vw,xy\in E(G)$ and such that $p\in (vw\cap xy)\setminus\{v,w,x,y\}$. An embedded graph $G$ is \emph{$k$-plane} if each edge of $G$ takes part in at most $k$ crossings.  A (not necessarily embedded) graph $G'$ is \emph{$k$-planar} if there exists $k$-plane graph $G$ isomorphic to $G'$.

@@

The remainder of the paper is organized as follows: In \secref{1-planar} we prove \thmref{1-planar}.  In \secref{k-planar} we prove \thmref{k-planar}.

@@

>>>>

\begin{abstract}
  Layered $H$-partitions of graphs (with small layered width, in which $H$ has small treewidth) are a recently-introduced tool that have been used to solve longstanding problems on queue layouts, non-repetitive colouring, and 3-d graph drawing.  Such partitions are known to exist for planar graphs and, more generally, bounded genus graphs.  In the current paper, we prove that every $k$-planar graph has a layered $H$-partition of layered width $O(k^2)$ in which $H$ has treewidth $O(c^k)$, for some constant $c$. This is the first result of this type for a non-minor-closed class of graphs and implies that $k$-planar graphs have both queue number and non-repetitive chromatic number upper-bounded by a function of $k$.  The former result was previously shown using a combination of $H$-partitions for planar graphs with \textit{ad hoc} methods. The latter result is new, for all $k\ge 1$.  These results extend to $(g,k)$-planar graphs, the natural generalization of $k$-planar graphs to to genus-$g$ surfaces (rather than the genus-$0$ plane).
\end{abstract}

@@

In this paper we show that another family of graphs admits layered $H$-partitions of constant layered-width in which $H$ has constant treewidth: $k$-planar graphs, which we now define. An \emph{embedded graph} $G$ is a graph with $V(G)\subset\R^2$ in which each edge $vw\in E(G)$ is a closed curve\footnote{A closed curve in a surface $S$ is a continuous function $f:[0,1]\to S$. The points $f(0)$ and $f(1)$ are called the \emph{endpoints} of the curve.  When there is no danger of misunderstanding we treat a curve $f$ as a point set $\{f(t):0\le t\le 1\}$.} in $\R^2$ with endpoints $v$ and $w$ and not containing any vertex of $G$ in its interior.  A \emph{crossing} in an embedded graph $G$ is a triple $(p,vw,xy)$ with $p\in\R^2$, $vw,xy\in E(G)$ and such that $p\in (vw\cap xy)\setminus\{v,w,x,y\}$. An embedded graph $G$ is \emph{$k$-plane} if each edge of $G$ takes part in at most $k$ crossings.  A (not necessarily embedded) graph $G'$ is \emph{$k$-planar} if there exists $k$-plane graph $G$ isomorphic to $G'$.

@@

There is a natural generalization of $k$-planar to genus-$g$ surfaces, that we now describe.  A graph $G$ embedded on a surface $S$ is $(S,k)$-plane if no edge of $G$ is involved in more than $k$ crossings.  A (not necessarily embedded) graph $G$ is $(g,k)$-planar if it is isomorphic to some $(S,k)$-plane graph $G'$ where $S$ is a surface with Euler genus $g$.

\begin{thm}\thmlabel{g-k-planar}
  Every $(g,k)$-planar graph has a layered $H$-partition of layered-width at most $\max\{16gk^2 + 24gk + 8g, 24k^2 + 36k + 12\}$ in which $H$ has treewidth at most $(3^{k+2}-3)/2$.

  Every $(g,1)$-planar graph has a layered $H$-partition of layered-width at most $\max\{48g, 30\}$ in which $H$ has treewidth at most $4$.  
\end{thm}

The remainder of the paper is organized as follows: In \secref{k-planar} we prove \thmref{k-planar}.  In \secref{1-planar} we prove \thmref{1-planar}.  
In \secref{genus-g} we prove \thmref{g-k-planar}.

@@

\section{$(g,k)$-Planar Graphs}
\seclabel{g-k-planar}

\begin{lem} \lemlabel{make-planar}
Let $G$ be a $(S,0)$-plane graph where $S$ is a surface with Euler genus $g$. 
%For every BFS layering $(V_0,V_1,\dots)$ of $G$,
%there is a set $Z\subseteq V(G)$ with at most $2g$ vertices in each layer $V_i$, such that $G-Z$ is planar. 
%Moreover, there is a connected planar graph $G^+$ containing $G-Z$ as a subgraph, 
%such that there is a BFS layering $(W_0,W_1,\dots)$ of $G^+$ and $W_{i} \cap V(G-Z) = V_i  \setminus Z$ for all $i\geq 0$. 
For every BFS spanning tree $T$ of $G$ rooted at some vertex $r\in V(G)$ with corresponding BFS layering $(V_0,V_1,\dots)$, 
there is a subgraph $Z\subseteq G$ with at most $2g$ vertices in each layer $V_i$, such that $G-V(Z)$ is planar. 
Moreover, there is a connected planar graph $G^*$ containing $G-V(Z)$ as a subgraph, and 
there is a BFS spanning tree $T^*$ of $G^*$ rooted at some vertex $r^*$ 
with corresponding BFS layering $(W_0,W_1,\dots)$ of $G^*$, 
such that  $W_{i} \cap V(G) \setminus V(Z) = V_i  \setminus V(Z)$ for all $i\geq 0$.
% , and
% $P \cap V(G) \setminus V(Z)$ is a vertical path in $T$ for every vertical path $P$ in $T^+$. 
\end{lem}

\begin{proof}[Proof of \thmref{g-k-planar}]
  Let $G$ be an $(S,k)$-plane graph where $S$ is a surface of Euler genus $g$.  Let $G^+$ be the $(S,0)$-plane graph obtained by adding a dummy vertex at each crossing in $G$.  Let $r$ be any vertex in $V(G)$ and let $T$ be BFS-spanning tree of $G^+$ rooted at $r$ and let $\mathcal{L}=\rangle V_0,ldots,V_h\rangle$ be the resulting layering of $G^+$.

  Apply \lemref{make-planar} to $G^{++}$ to obtain the set $Z\subseteq V(G^{++})$, the planar graph $G^*$, and the BFS spanning tree $T^*$ rooted at $r^*$.  For each edge $vw\in E(G)$, let $W_{vw}$ be the corresponding walk in $G^+$. The set $Z$ contains vertices in $V(G)$ as well as dummy vertices in $V(G^+-G)$. Create a new set $Z'\subseteq V(G)$ as
  \[
     Z' = \{ v : V(W_{vw})\cap Z\neq\emptyset \} \enspace .
  \]
  Since $|Z\cap V_i| \le 2g$, the same argument used in the proof of \lemref{width-of-g-plus} shows that $|Z'\cap V_i|\le 2g\times 4\times (2k+1) = 16gk+8g$ for all $i\in\{1,\ldots,h\}$.

  Furthermore, for every edge $vw\in E(G-Z')$, $\dist_{G^+-Z}(v,w)\le k+1$, since $W_{vw}\subseteq G^+-Z$.

  Next, observe that any plane embedding of $G^*\supseteq G^+-Z$ gives a $k$-plane embedding of $G-Z'$. Indeed, for every edge $vw\in E(G-Z')$, $G^+-Z$ contains a walk $W_{vw}$ that contains at most $k$ dummy vertices of $G^+-Z$.  Therefore, replacing edge edge $vw\in E(G-Z')$ with the corresponding walk in $G^*$ gives a $k$-plane embedding of $G-Z'$.

  Apply \thmref{k-planar} or (in the case $k=1$) \thmref{1-planar} to the $k$-plane embedding of $G-Z'$ to obtain a layered $H$-partition $(\mathcal{L}',\mathcal{P})$ of $G-Z'$.  Though not explicitly in the statement of either theorem, the proof of each theorem shows that it is possible to choose a vertex $r^*\in G^*$ so that each layer $V'_i\in \mathcal{L}'$ is contained in at most $k+1$ consecutive layers of $\mathcal{L}$.  Therefore, $|Z'\cap V_i'| \le (k+1)(16gk+8g) = 16gk^2 + 24gk + 8g$.

  We turn $\mathcal{P}$, the $H$-partition of $G-Z'$ into a partition $\mathcal{P}'=\mathcal{P}\cup \{Z'\}$.  Clearly, the treewidth of $H'=G/\mathcal{P'}$ is at most one more than the treewidth of $H=G/\mathcal{P}$.  Finally
  \[
      \max\{ |P\cap L| : P\in \mathcal{P}',\, L\in \mathcal{L}' \} \le
      \max\{ 16gk^2 + 24gk + 8g, 24k^2 + 36k + 12 \} \enspace .
  \]
  \note{PM}{All of the above still needs work...}
\end{proof}
